A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-22 paper/paper.tex
paper: minor nits

%TODO check in other places if it makes sense to use the notation and definition LaTeX environment
\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}      \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}                             % TODO: n not bound on the left
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as what is called the Yoneda embedding of $n\in\Obj(\DeltaPlus)$:

@@

More generally, the standard augmented $(n+1)$-semi-simplex can be obtained by taking a copy of the standard augmented $n$-semi-simplex serving as the base and glueing on top of it another copy of it lifted by one dimension. In the latter, the color becomes an extra point, the points become lines connecting the points of the base to the extra point and so on. In particular, the components of the base are those of the standard augmented $n$-semi-simplex postfixed by $0$ while the components of the lifted copy are postfixed by $\kstar$.

@@

\begin{definition}[$\Cube^n$]
  The standard semi-cube $\Cube^n$ is defined as the Yoneda embedding of $n\in\Obj(\Cube)$:

@@

More generally, the standard $(n+1)$-semi-cube can be obtained by taking two copies of the standard $n$-semi-cube serving as bottom and top face and connecting them by a tube obtained as a third copy stretched in the new dimension. The bottom and top faces are obtained from the standard $n$-semi-cube by postfixing with respectively $L$ and $R$ while the tube is obtained by postfixing with~$\kstar$.

@@

Now, one may wonder how to express the type of $X_i$ uniformly as a function of $n$? Connecting semi-cubical sets to binary parametricity
%, later, thanks to the generalization to $\nu$-sets, augmented semi-simplicial sets to unary parametricity
will give us a solution which we describe in the next section.

@@

\subsection{Relating the indexed representation to parametricity}

The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes two types $A_L$ and $A_R$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$: %TODO be explicit that HSet and HSet* are in the next universe??

@@

\begin{equation*}
  \begin{array}{llllr}
    X_0 & \defeq & X                & : & \U                                               \\
    X_1 & \defeq & X_\kstar         & : & {\U}_\kstar(X)(X)                                \\
    X_2 & \defeq & X_{\kstar\kstar} & : & ({\U}_\kstar(X)(X))_{\kstar}(X_\kstar)(X_\kstar) \\
    \ldots
  \end{array}
\end{equation*}
This tells us how the informal type given to $X_2$ in the previous section could be rephrased so that it comes as the instance of a general recipe characterizing the type of all $X_i$.

@@

We illustrated here the cubical case, that is $\nu=2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\filler[n - 1][p]$. A $\filler[n][p]$ is an $n$-dimensional object obtained by stretching a $(n - p)$-dimensional frame in $p$ directions. For instance, in the picture, each of the two $\filler[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\filler[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a square, and each of the two $\filler[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a square. Note that a filler, which extends a frame with an element depending on it, shown as an element of $X_n$ below. As a filler extends a frame, it also has layer components, but note that the layers of a filler are associated on the right that is opposite to the way they are associated in frames:
\begin{align*}
  \framep[n][p] & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0]) \ldots) .\; \layer[n][p] \\
  \filler[n][p] & \defeq \Sigma l_p : \layer[n][p] . (\ldots (\Sigma l_n : \layer[n][n - 1] .\; X_n) \ldots)
\end{align*}

>>>>

%TODO check in other places if it makes sense to use the notation and definition LaTeX environment
\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}      \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as what is called the Yoneda embedding of $n \in \Obj(\DeltaPlus)$:

@@

More generally, the standard augmented $(n+1)$-semi-simplex can be obtained by taking a copy of the standard augmented $n$-semi-simplex serving as the base and glueing on top of it another copy of it lifted by one dimension. In the second copy, the color becomes an extra point, the points become lines connecting the points of the base to the extra point and so on. In particular, the components of the base are those of the standard augmented $n$-semi-simplex postfixed by $0$ while the components of the lifted copy are postfixed by $\kstar$.

@@

\begin{definition}[$\Cube^n$]
  The standard semi-cube $\Cube^n$ is defined as the Yoneda embedding of $n \in \Obj(\Cube)$:

@@

More generally, the standard $(n+1)$-semi-cube can be obtained by taking two copies of the standard $n$-semi-cube serving as bottom and top face and connecting them by a cylinder obtained as a third copy stretched in the new dimension. The bottom and top faces are obtained from the standard $n$-semi-cube by postfixing with respectively $L$ and $R$ while the cylinder is obtained by postfixing with \kstar.

@@

Now, one may wonder how to express the type of $X_i$ uniformly as a function of $n$. Connecting semi-cubical sets to binary parametricity will give us a solution which we describe in the next section.

@@

\subsection{Relating the indexed representation to parametricity}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:

@@

\begin{equation*}
  \begin{array}{llllr}
    X_0 & \defeq & X                & : & \U                                               \\
    X_1 & \defeq & X_\kstar         & : & {\U}_\kstar(X)(X)                                \\
    X_2 & \defeq & X_{\kstar\kstar} & : & ({\U}_\kstar(X)(X))_{\kstar}(X_\kstar)(X_\kstar) \\
    \ldots
  \end{array}
\end{equation*}

This tells us how the informal type given to $X_2$ in the previous section could be rephrased so that it comes as the instance of a general recipe characterizing the type of all $X_i$.

@@

We illustrated here the cubical case, that is $\nu = 2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\filler[n - 1][p]$. A $\filler[n][p]$ is an $n$-dimensional object obtained by stretching a $(n - p)$-dimensional frame in $p$ directions. For instance, in the picture, each of the two $\filler[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\filler[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a square, and each of the two $\filler[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a square. Note that a filler, which extends a frame with an element depending on it, shown as an element of $X_n$ below. As a filler extends a frame, it also has layer components, but note that the layers of a filler are associated on the right that is opposite to the way they are associated in frames:

\begin{align*}
  \framep[n][p] & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0]) \ldots) .\; \layer[n][p] \\
  \filler[n][p] & \defeq \Sigma l_p : \layer[n][p] . (\ldots (\Sigma l_n : \layer[n][n - 1] .\; X_n) \ldots)
\end{align*}
